---
title: "Assignment 4 Task 2"
author: "Logan Babcock"
date: "November 4, 2021"
output: html_document
---

```{r setup, include=FALSE, messages = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE,  messages = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(janitor)
library(kableExtra)
library(lubridate)


hares_raw <- read.csv(here("data", "bonanza_hares.csv"))

```

## 1. Overview
This is an exploratory report on juvenile snoeshoe hares (*Lepus americanus*) using data collected during the  US Long Term Ecological Research Network's study on Snowshoe hare physical data in the Bonanza Creek Experimental Forest. Using the data collected from 1999-2012, this report will analyze the counts and sizes of juvenile hares.

*COME BACK TO FINISH THIS  - add info on what the report contains*

## 2. Data and Methods
Snowshoe hare data and measurements were made accessible through Knut Kielland and associates at the Bonanza Creek LTER. The data contain measurements and observations on hares during a capture-recapture study done in Alaska's Tanana valley from 1999-2012. 

*COME BACK TO FINISH THIS - add info on stats tests and whats included* 

## Data Visualization and Analysis
```{r}

# finding juvenile hare counts

# making separate year column
hares_year <- hares_raw %>% # making new dataset with new date column
  mutate(date_new = mdy(date)) %>% # new column with date class
  mutate(year = year(date_new)) # new column with just year

yearly_count <- hares_year %>% # making a dataset wiht just year and count
  group_by(year) %>% # grouping by year
  count() %>%  #adding all observations for that year
  rename(Year = year, Count = n)

kable(yearly_count, caption = "Table 1: Number of juvenlie snowshoe hares caught per year" , align = "l") %>% # making a finalized table with title
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "left", wraptable_width = 0) # customizing table

# visualizing the data

ggplot(data = yearly_count, aes(x = Year, y = Count)) +
  geom_col(color = "coral2", fill = "coral2") +
  scale_x_continuous(limits=c(1997,2013), breaks = c(1998, 2000, 2002, 2004, 2006, 2008, 2010, 2012)) +
  labs(caption = "Figure 1: Number of juvenile hares caught during sampling from 1998-2012, shown through bar chart with height representing number of individual hares.", title = "Hares caught each year (Babcock)")
```
Above, Figure 1 shows the number of hares caught each year of the survey period. The number of hares caught vaired between 38 and 410. 